---
title: "A Quick Latent Class Analysis (LCA) from Start to Finish in `MplusAutomation`"
author: '*Adam Garber*'
subtitle: "IMMERSE Video Series Funded by IES "
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

What is included in this video tutorial?

A demonstration of the speed at which an LCA analysis can be estimated and summarized using the `Tidy MplusAutomation` method.  

# --------------------------------------------------------------------------------------

## Tutorial Outline
   
1.  
2.  
3.  Install & load packages
4.  Read in data to R
5.  Look at LCA indicator variables
6.  Run a `type=basic` model using `MplusAutomation`
7.  
8.  
9.  

# -------------------------------------------------------------------------------------- 

## Github repository (everything you need to replicate analysis):

https://github.com/garberadamc/QF-Getting-Started


# -------------------------------------------------------------------------------------- 

## Load packages

# --------------------------------------------------------------------------------------

```{r,eval=TRUE}

library(MplusAutomation); library(glue) # estimation
library(tidyverse); library(here); # tidyness 
library(gt); library(reshape2); library(cowplot) # tables & figures

```

# --------------------------------------------------------------------------------------

Data Source Information:

https://github.com/UrbanInstitute/education-data-package-r

https://educationdata.urban.org/documentation/index.html#data_sources

Variable terms & definitions
https://www.isbe.net/Documents/Civil-Rights-Data-Collection-Terms.pdf

# --------------------------------------------------------------------------------------

## Read in CSV data file from the `data` subfolder

```{r}

bully_data <- read_csv(here("data", "crdc_lca_data.csv"))

```

# --------------------------------------------------------------------------------------

## Indicator Variables

`report_dis`  = Number of students harassed or bullied on the basis of disability 
`report_race` = Number of students harassed or bullied on the basis of race, color, or national origin  
`report_sex`  = Number of students harassed or bullied on the basis of sex 
`counselors_fte` = Number of full time equivalent counselors hired as school staff
`psych_fte` = Number of full time equivalent psychologists hired as school staff
`law_fte` = Number of full time equivalent law enforcement officers hired as school staff


# --------------------------------------------------------------------------------------

## Run a first model using the `mplusObject()` method 

Model is `type = BASIC;` (i.e., returns descriptive statistics)
```{r,eval=FALSE}

m_basic  <- mplusObject(
  
  TITLE = "Explore descriptives with TYPE = BASIC", 
  
  VARIABLE = 
 "",
  
  ANALYSIS = 
 "",
 
  usevariables = colnames(), 
  rdata = )

m_basic_fit <- mplusModeler( , 
               dataout=here("", ""),
               modelout=here("", ".inp"),
               check=TRUE, run = TRUE, hashfilename = FALSE)

```

## Always check your model!

- In the RStudio window pane on the `bottom-right`under the `files` tab click on the `mplus_files` folder
- There should be 3 new files in this location that were produced by `mplusModeler()`
- Click on the output file (`.out`) to check if the model estimated or if there are any error messages

# --------------------------------------------------------------------------------------

# Enumeration

# --------------------------------------------------------------------------------------

```{r, eval=FALSE}

lca_k1_8  <- lapply(1:8, function(k) {
  
  lca_enum  <- mplusObject(
      
    TITLE = glue("Class {k} LCA"), 
  
    VARIABLE = glue(
    "categorical = ; 
     usevar = ;
     classes = c({k}); "),
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = ; 
    processors = 10;",
  
  OUTPUT = "tech11 tech14;",
  
  PLOT = 
    "type = plot3; 
     series = (*);",
  
  usevariables = colnames(bully_staff),
  rdata = bully_staff)

lca_enum_fit <- mplusModeler(lca_enum, 
                            dataout=glue(here("", ".dat")),
                            modelout=glue(here("", "c{k}_.inp")) ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
})

```

# --------------------------------------------------------------------------------------

##  1.2 Generate Model Fit Summary Table 

- This syntax can be used to compare model fit from the series of LCA models generated during enumeration
- The code produces a table that is approximately in APA format. 

# --------------------------------------------------------------------------------------

Read in model fit statistics using `readModels()` and `mixtureSummaryTable()` functions
```{r}
output_lca <- readModels(here(""), quiet = TRUE)

enum_summary <- LatexSummaryTable(output_lca,                                          
                keepCols=c("Title", "Parameters", "LL", "BIC", "aBIC",
                           "BLRT_PValue", "T11_VLMR_PValue","Observations"), 
                sortBy = "Title")
```

Calculate relevant fit indices for summary table
```{r}
allFit <- enum_summary %>% 
  mutate(aBIC = -2*LL+Parameters*log((Observations+2)/24)) %>% 
  mutate(CIAC = -2*LL+Parameters*(log(Observations)+1)) %>% 
  mutate(AWE = -2*LL+2*Parameters*(log(Observations)+1.5)) %>%
  mutate(SIC = -.5*BIC) %>% 
  mutate(expSIC = exp(SIC - max(SIC))) %>% 
  mutate(BF = exp(SIC-lead(SIC))) %>% 
  mutate(cmPk = expSIC/sum(expSIC)) %>% 
  dplyr::select(1:5,9:10,6:7,13,14) %>% 
  arrange(Parameters)
```


Generate the fit summary table
```{r}
allFit %>% 
  mutate(Title = str_remove(Title, " LCA Enumeration ")) %>% 
  gt() %>%
  tab_header(
    title = md("**Model Fit Summary Table**"), subtitle = md("&nbsp;")) %>% 
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    BF = md("BF"),
    cmPk = md("*cmPk*")) %>%
  tab_footnote(
    footnote = md(
    "*Note.* Par = parameters; *LL* = log likelihood;
      BIC = bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      BLRT = bootstrapped likelihood ratio test p-value;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
      cmPk = approximate correct model probability."), 
    locations = cells_title()) %>% 
  tab_options(column_labels.font.weight = "bold") %>% 
  fmt_number(10,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt_number(c(3:9,11), 
             decimals = 0) %>% 
  fmt_missing(1:11,
              missing_text = "--") %>% 
  fmt(c(8:9,11),
    fns = function(x) 
    ifelse(x<0.001, "<.001",
           scales::number(x, accuracy = 0.01))) %>%
  fmt(10, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) 

```


# --------------------------------------------------------------------------------------

##  1.3 Plot Information Criteria 

# --------------------------------------------------------------------------------------

```{r}
allFit %>% dplyr::select(2:7) %>% 
  rowid_to_column() %>% 
  pivot_longer(`BIC`:`AWE`, 
    names_to = "Index", 
    values_to = "ic_value") %>% 
  mutate(Index = factor(Index,
    levels = c("AWE","CIAC","BIC","aBIC"))) %>%
  ggplot(aes(x = rowid, y = ic_value,
    color = Index, shape = Index,
    group = Index, lty = Index)) + 
  geom_point(size = 2.0) + geom_line(size = .8) +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Number of Classes", y = "Information Criteria Value") +
  theme_cowplot() + theme(legend.title = element_blank(), legend.position = "top")
```
```{r, eval=FALSE}
ggsave(here("figures","FIT_BullyStaff_TX17.png"),    
       dpi=300, height=4, width=6, units="in")    
```

# --------------------------------------------------------------------------------------

##  1.4 Compare Conditional Item Probability Plots

# --------------------------------------------------------------------------------------

```{r, fig.width = 7, fig.height = 5}

model_results <- data.frame()
for (i in 1:length(output_17)) {
  temp <- data.frame(unclass(output_17[[i]]$parameters$unstandardized)) %>% 
    mutate(model = paste0(i, "-Class Model")) 
  model_results <- rbind(model_results, temp) }

pp_plots <- model_results %>% 
  filter(paramHeader == "Thresholds") %>% dplyr::select(est, model, LatentClass, param) %>% 
  mutate(prob = (1 / (1 + exp(est))), param = str_to_title(str_remove_all(param, "[$1]")))

ggplot(pp_plots,
       aes(x = param, y = prob, color = LatentClass, shape = LatentClass, group = LatentClass)) + 
  geom_point() + geom_line() + facet_wrap(~ model, ncol = 2) + labs(x= "", y = "Probability") +
  theme_minimal() + theme(legend.position = "none", axis.text.x = element_text(size = 6))
```


# --------------------------------------------------------------------------------------

##  1.5 Plot Final Model - Conditional Item Probability Plot 

# --------------------------------------------------------------------------------------

### This syntax creates a function called `plot_lca_function` that requires 7 arguments (inputs):

- `model_name`: name of Mplus model object (e.g., `model_step1`)
- `item_num`: the number of items in LCA measurement model (e.g., `5`)
- `class_num`: the number of classes (*k*) in LCA model (e.g., `3`)
- `item_labels`: the item labels for x-axis (e.g., `c("Enjoy","Useful","Logical","Job","Adult")`)
- `class_labels`: the class label names (e.g., `c("Adaptive Coping","Externalizing Behavior","No Coping")`)
- `class_legend_order` = change the order that class names are listed in the plot legend (e.g., `c(2,1,3)`)
- `plot_title`: include the title of the plot here (e.g., `"LCA Posterior Probability Plot"`)

Read in plot data from Mplus output file `c3_lca.out`
```{r}

model_c3 <- readModels(here("enum2_CA17", "c3_lca.out"), quiet = TRUE)
                           
```

```{r}

plot_lca_function <- function(model_name,item_num,class_num,item_labels,
                              class_labels,class_legend_order,plot_title){

mplus_model <- as.data.frame(model_name$gh5$means_and_variances_data$estimated_probs$values)
plot_data <- mplus_model[seq(2, 2*item_num, 2),]

c_size <- as.data.frame(model_name$class_counts$modelEstimated$proportion)
colnames(c_size) <- paste0("cs")
c_size <- c_size %>% mutate(cs = round(cs*100, 2))
colnames(plot_data) <- paste0(class_labels, glue(" ({c_size[1:class_num,]}%)"))
plot_data <- plot_data %>% relocate(class_legend_order)

plot_data <- cbind(Var = paste0("U", 1:item_num), plot_data)
plot_data$Var <- factor(plot_data$Var,
               labels = item_labels)
plot_data$Var <- fct_inorder(plot_data$Var)

pd_long_data <- melt(plot_data, id.vars = "Var") 

# This syntax uses the data.frame created above to produce the plot with `ggplot()`

p <- pd_long_data %>%
  ggplot(aes(x = as.integer(Var), y = value,
  shape = variable, colour = variable, lty = variable)) +
  geom_point(size = 4) + geom_line() + 
  scale_x_continuous("", breaks = 1:item_num,
                     labels = function(x) str_wrap(plot_data$Var, width = 13)) + 
  labs(title = plot_title, y = "Probability") +
  theme_cowplot() +
  theme(legend.title = element_blank(), 
        legend.position = "top",
        axis.text.x = element_text(size=8))

p
return(p)
}

```

## Run C3 Plot
```{r}

plot_lca_function(
  model_name = model_c3, 
  item_num = 6,
  class_num = 3,
  item_labels = c("harassment: disability","harassment: race","harassment: sex",
                  "school staff: counselor","school staff: psychologist","school staff: law enforcement"),
  class_labels = c("C1","C2","C3"),
  class_legend_order = c(1,2,3),
  plot_title = "California: Harrasment & School Staff (K = 3)"
  )

```

```{r, eval=FALSE}
ggsave(here("",".png"),    
       dpi=300, height=4, width=6, units="in")    
```




